---
<<<<<<< HEAD
title: "Project 4: Baysian modeling of hurrican trajectories"
=======
title: 'Project 3: Bayesian modeling of hurricane trajectories'
>>>>>>> 198e79a4f5745ebd7448378ee32e24267cc86359
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---


# Hurricane Data
hurricane356.csv collected the track data of 356 hurricanes in the North Atlantic area since 1989. For all the storms, their location (longitude \& latitude) and maximum wind speed were recorded every 6 hours. The data includes the following variables 

1. **ID**: ID of the hurricanes
2. **Season**: In which \textbf{year} the hurricane occurred 
3. **Month**: In which \textbf{month} the hurricane occurred 
4. **Nature**: Nature of the hurricane 
  + ET: Extra Tropical
  + DS: Disturbance
  + NR: Not Rated
  + SS: Sub Tropical
  + TS: Tropical Storm
5. **time**: dates and time of the record  
6. **Latitude** and **Longitude**: The location of a hurricane check point 
7. **Wind.kt** Maximum wind speed (in Knot) at each check point 



## Load and explore the hurricane data through visulaizations

```{r eval=FALSE, include=TRUE}
library(ggplot2)
dt= read.csv("/cloud/project/hurrican356.csv")
ggplot(data=dt, aes(x = Longitude, y = Latitude)) + 
  stat_summary_2d(data = dt, aes(x = Longitude, y = Latitude, z = dt$Wind.kt), fun = median, binwidth = c(1, 1), show.legend = TRUE)
library(data.table)
dt <- as.data.table(dt)
summary(dt)
```
Overlay the hurricane data in the world map
```{r eval=FALSE, include=TRUE}
library(maps)
map <- ggplot(data = dt, aes(x = Longitude, y = Latitude)) + 
  geom_polygon(data = map_data(map = 'world'), aes(x = long, y = lat, group = group))
map +
  stat_summary_2d(data = dt, aes(x = Longitude, y = Latitude, z = dt$Wind.kt), fun = median, binwidth = c(1, 1), show.legend = TRUE, alpha = 0.75) + 
  ggtitle(paste0("Atlantic Windstorm mean knot"))
```
Additional Plots
```{r eval=FALSE, include=TRUE}
map <- ggplot(dt, aes(x = Longitude, y = Latitude, group = ID)) + 
  geom_polygon(data = map_data("world"), 
               aes(x = long, y = lat, group = group), 
               fill = "gray25", colour = "gray10", size = 0.2) + 
  geom_path(data = dt, aes(group = ID, colour = Wind.kt), size = 0.5) + 
  xlim(-138, -20) + ylim(3, 55) + 
  labs(x = "", y = "", colour = "Wind \n(knots)") + 
  theme(panel.background = element_rect(fill = "gray10", colour = "gray30"),
        axis.text.x = element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

seasonrange <- paste(range(dt[, "Season"]), collapse=" - ")

map + ggtitle(paste("Atlantic named Windstorm Trajectories (", 
                     seasonrange, ")\n")) 
```
Show hurricane tracks by month
```{r eval=FALSE, include=TRUE}
mapMonth <- map + facet_wrap(~ Month) +
  ggtitle(paste("Atlantic named Windstorm Trajectories by Month (", 
                 seasonrange, ")\n")) 
mapMonth
```



# A Hierarchical Bayesian model for hurricane trajectories. 

Climate researchers are interested in modeling the hurricane trajectories to forecast the wind speed. Let $t$ be time (in hours) since a hurricane began, and For each hurricane $i$, we denote $Y_{i}(t)$ be the wind speed of the $i$th hurricane at time $t$. The following Bayesian model was suggested.  


$$Y_{i}(t+6) =\beta_{0,i}+x_{i,1}\beta_{1,i} +
<<<<<<< HEAD
x_{i,2} \beta_{2,i} + x_{i,3}\beta_{3,i} +\beta_{4,i,j}Y_{i,j}(t) +  +\beta_{5,i,j}\Delta_{i,1}(t)+ +\beta_{6,i,j}\Delta_{i,2}(t)+ +\beta_{7,i}\Delta_{i,3} + \epsilon_{i}(t)$$   
where $x_{i,1}$ is the month of year when the hurricane started, $x_{i,2}$ is the calendar year of the hurricane, and $x_{i,3}$ is the type of hurricane, $\Delta_{i,1}(t)$, $\Delta_{i,2}(t)$ and $\Delta_{i,3}(t)$ is the change of latitude longitude, and wind speed between $t-6$ and $t$, and $\epsilon_{i,t}$ follows a  normal distributions with mean zero and variance $\sigma^2$, independent across $t$.
=======
x_{i,2} \beta_{2,i} + x_{i,3}\beta_{3,i} +\beta_{4,i}Y_{i}(t) +\beta_{5,i}\Delta_{i,1}(t)+ \beta_{6,i}\Delta_{i,2}(t)+ +\beta_{7,i}\Delta_{i,3} + \epsilon_{i}(t)$$

where $x_{i,1}$ is the month of year when the hurricane started, $x_{i,2}$ is the calendar year of the hurricane, and $x_{i,3}$ is the type of hurricane, $\Delta_{i,1}(t)$, $\Delta_{i,2}(t)$ and $\Delta_{i,3}(t)$ is the change of latitude longitude, and wind speed between $t-6$ and $t$, and $\epsilon_{i,t}$ follows a normal distributions with mean zero and variance $\sigma^2$, independent across $t$.
>>>>>>> 198e79a4f5745ebd7448378ee32e24267cc86359


In the model, $\boldsymbol{\beta}_{i} = (\beta_{0,i},\beta_{1,i},...,\beta_{7,i})$ are the random coefficients associated the $i$th hurricane, we assume that 

$$\boldsymbol{\beta}_{i} \sim N(\boldsymbol{\beta}, \boldsymbol{\Sigma})$$
follows a multivariate normal distributions with mean $\boldsymbol{\beta}$ and covariance matrix $\Sigma$.


\paragraph{Prior distributions}


We assume the following non-informative or weak prior distributions for $\sigma^2$, $\boldsymbol{\beta}$ and $\Sigma$.
$$P(\sigma^2) \propto \frac{1}{\sigma^2};\quad P(\boldsymbol{\beta})\propto 1;\quad P(\Sigma^{-1}) \propto 
|\Sigma|^{-(d+1)} \exp(-\frac{1}{2}\Sigma^{-1})$$
d is dimension of $\beta$.


## Design a MCMC algorithm to estiamte the posterior distributions of the model parameters, and complete the following  tasks:

1. Construct their 95% credibility intervals for $\boldsymbol{\beta}$. What we learn from the models? Is there evidence support that the statement that "the hurricane wind speed has been increasing over years". 

2. How well does this Bayesian model track the individual hurricanes?  


3. Write a summary statement of your findings and comment on the proposed Bayesian model.
