---
title: 'Project 3: Bayesian modeling of hurricane trajectories'
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---


# Hurricane Data
hurricane356.csv collected the track data of 356 hurricanes in the North Atlantic area since 1989. For all the storms, their location (longitude \& latitude) and maximum wind speed were recorded every 6 hours. The data includes the following variables 

1. **ID**: ID of the hurricanes
2. **Season**: In which \textbf{year} the hurricane occurred 
3. **Month**: In which \textbf{month} the hurricane occurred 
4. **Nature**: Nature of the hurricane 
  + ET: Extra Tropical
  + DS: Disturbance
  + NR: Not Rated
  + SS: Sub Tropical
  + TS: Tropical Storm
5. **time**: dates and time of the record  
6. **Latitude** and **Longitude**: The location of a hurricane check point 
7. **Wind.kt** Maximum wind speed (in Knot) at each check point 



## Load and explore the hurricane data through visulaizations

```{r eval=FALSE, include=TRUE}
library(ggplot2)
# dt= read.csv("/cloud/project/hurrican356.csv")
dt= read.csv("./hurrican356.csv")
ggplot(data=dt, aes(x = Longitude, y = Latitude)) + 
  stat_summary_2d(data = dt, aes(x = Longitude, y = Latitude, z = dt$Wind.kt), fun = median, binwidth = c(1, 1), show.legend = TRUE)
library(data.table)
dt <- as.data.table(dt)
summary(dt)
```

Overlay the hurricane data in the world map
```{r eval=FALSE, include=TRUE}
library(maps)
map <- ggplot(data = dt, aes(x = Longitude, y = Latitude)) + 
  geom_polygon(data = map_data(map = 'world'), aes(x = long, y = lat, group = group))
map +
  stat_summary_2d(data = dt, aes(x = Longitude, y = Latitude, z = dt$Wind.kt), fun = median, binwidth = c(1, 1), show.legend = TRUE, alpha = 0.75) + 
  ggtitle(paste0("Atlantic Windstorm mean knot"))
```

Additional Plots
```{r eval=FALSE, include=TRUE}
map <- ggplot(dt, aes(x = Longitude, y = Latitude, group = ID)) + 
  geom_polygon(data = map_data("world"), 
               aes(x = long, y = lat, group = group), 
               fill = "gray25", colour = "gray10", size = 0.2) + 
  geom_path(data = dt, aes(group = ID, colour = Wind.kt), size = 0.5) + 
  xlim(-138, -20) + ylim(3, 55) + 
  labs(x = "", y = "", colour = "Wind \n(knots)") + 
  theme(panel.background = element_rect(fill = "gray10", colour = "gray30"),
        axis.text.x = element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

seasonrange <- paste(range(dt[, "Season"]), collapse=" - ")

map + ggtitle(paste("Atlantic named Windstorm Trajectories (", 
                     seasonrange, ")\n")) 
```

Show hurricane tracks by month
```{r eval=FALSE, include=TRUE}
mapMonth <- map + facet_wrap(~ Month) +
  ggtitle(paste("Atlantic named Windstorm Trajectories by Month (", 
                 seasonrange, ")\n")) 
mapMonth
```

# A Hierarchical Bayesian model for hurricane trajectories. 

Climate researchers are interested in modeling the hurricane trajectories to forecast the wind speed. Let $t$ be time (in hours) since a hurricane began, and For each hurricane $i$, we denote $Y_{i}(t)$ be the wind speed of the $i$th hurricane at time $t$. The following Bayesian model was suggested.  


$$Y_{i}(t+6) =\beta_{0,i}+x_{i,1}\beta_{1,i} +
x_{i,2} \beta_{2,i} + x_{i,3}\beta_{3,i} +\beta_{4,i}Y_{i}(t) +\beta_{5,i}\Delta_{i,1}(t)+ \beta_{6,i}\Delta_{i,2}(t)+ +\beta_{7,i}\Delta_{i,3} + \epsilon_{i}(t)$$

where $x_{i,1}$ is the month of year when the hurricane started, $x_{i,2}$ is the calendar year of the hurricane, and $x_{i,3}$ is the type of hurricane, $\Delta_{i,1}(t)$, $\Delta_{i,2}(t)$ and $\Delta_{i,3}(t)$ is the change of latitude longitude, and wind speed between $t-6$ and $t$, and $\epsilon_{i,t}$ follows a normal distributions with mean zero and variance $\sigma^2$, independent across $t$.


In the model, $\boldsymbol{\beta}_{i} = (\beta_{0,i},\beta_{1,i},...,\beta_{7,i})$ are the random coefficients associated the $i$th hurricane, we assume that 

$$\boldsymbol{\beta}_{i} \sim N(\boldsymbol{\beta}, \boldsymbol{\Sigma})$$
follows a multivariate normal distributions with mean $\boldsymbol{\beta}$ and covariance matrix $\Sigma$.


\paragraph{Prior distributions}


We assume the following non-informative or weak prior distributions for $\sigma^2$, $\boldsymbol{\beta}$ and $\Sigma$.
$$P(\sigma^2) \propto \frac{1}{\sigma^2};\quad P(\boldsymbol{\beta})\propto 1;\quad P(\Sigma^{-1}) \propto 
|\Sigma|^{-(d+1)} \exp(-\frac{1}{2}\Sigma^{-1})$$
d is dimension of $\beta$.


## Design a MCMC algorithm to estiamte the posterior distributions of the model parameters, and complete the following  tasks:

1. Construct their 95% credibility intervals for $\boldsymbol{\beta}$. What we learn from the models? Is there evidence support that the statement that "the hurricane wind speed has been increasing over years". 

2. How well does this Bayesian model track the individual hurricanes?  


3. Write a summary statement of your findings and comment on the proposed Bayesian model.


# A Hierarchical Bayesian model for hurricane trajectories. 

## Introduction

  Hierarchical Bayesian Model is the combination of two methods: Linear Regression Model in Hierarchical Form and Bayesian Inference methods.
  
  Hierarchical form model is the linear regression model containing both Within-group analysis and Between-group analysis. Observations, from hierarchical form model, are usually collected with natural heterogeneity across the population of subjects over research time periods. This natural heterogeneity can be regarded as subject-specific mean response trajectories (i,e,. random effects) for each individual group. Holding these individual-specific effects, the overall mean response over time across the whole research population is still considered as linearly over time, which is called Population-level effects (i.e,. fixed effects).
  $$\underbrace{\pi(\theta| X)}_{posterior\ distribution} \propto \underbrace{\pi(X|\theta)}_{likelihood} \times \underbrace{\pi(\theta)}_{prior\ distribution} $$
  Bayesian Inference is a statistical inference method about parameter. Before data collection, a proper prior distribution of parameter $\theta$ is set based on our belief about $\theta$. Then after data collection $\textbf{X} = (X_1, X_2,..., X_n)$, the belief of parameter $\theta$ would be updated by exploring the posterior distribution of $\theta$ based on observed data and its pre-assumed likelihood function $L(X; \theta)$. The  linear regression model in hierarchical form incorporating with Bayesian inference is implemented with Markov Chain Monte Carlo Integration algorithm for updating the parameter estimation in the final MCMC stationary phase. 
  
  In this project, we are going to explore whether the population-level changing trend of hurricane wind speed over years through a total of `r length(table(data$ID))` groups of hurricanes. Each hurricane contained its own individual-level-specific effects. The hierarchical Bayesian model for the $\textit{i}$th hurricane is shown as
$$Y_i(t + 6) = \beta_{0,i} + X_{i_{Month}}\beta_{1,i} + X_{i_{Year}}\beta_{2,i} + X_{i_{Type}}\beta_{3,i} + X_{i_{Y_{i,t}(t)}}\beta_{4,i,t} + \Delta_{i,t_{lat}}\beta_{5,i,t} + \Delta_{i,t_{lon}}\beta_{6,i,t} + \Delta_{i,t_{Speed}}\beta_{7,i,t} + \epsilon_i(t),$$ 
where $i = 1,\ 2,..., m$ standing for each hurricane group and $t = 1,\ 2,..., t_i$ standing for each recorded time point within $\textit{i}$th hurricane group.

  The provided hierarchical Bayesian model for hurricane trajectories for $\textit{i}$th hurricane contains 4 population-level effects (i.e,. fixed): $X_{i_{Month}}$ - the month of year when hurricane started,  $X_{i_{Year}}$ - the calender year of the hurricane, $X_{i_{Type}}$ - the type of hurricane, and 4 individual-level- effects (i.e,. random): $X_{i_{Y_{i,ti}(t)}}$ - the $\textit{i}$th wind speed at $t - 1$ time point for t time point, $\Delta_{i,t_{lat}},\ \Delta_{i,t_{lon}},\ \Delta_{i,t_{Speed}}$ - the change of latitudes, longitudes and wind speeds between two recorded time points. 4 prior information are provided as following:
$$\boldsymbol{\beta_i} \sim N(\boldsymbol{\beta}, \Sigma^{-1}),\ \pi(\sigma^2) \propto \frac{1}{\sigma^2},$$
$$\pi(\boldsymbol{\beta}) \propto 1,\ \pi(\boldsymbol{\Sigma}^{-1}) \propto |\boldsymbol{\Sigma|}^{-(d + 1)/2}exp(-\frac{1}{2}\boldsymbol{\Sigma}^{-1}),$$
  Then our Hierarchical Bayesian Model for analyzing the trend of hurricane wind speed across years with considering all other covariates would be performed firstly by exploring estimated parameters from posterior distributions based on Bayes Theorem. 
  

## Posterior Distribution Inference and MCMC Methods

### Posterior Distribution Inference
  
  We have $\beta_i = (\beta_{0,i}, \beta_{1,i},..., \beta_{7,i,t})$ associated with the $\textit{i}$th hurricane. According to the knowledge of Multivariate Linear Regression Model, we know that $Y_{i,t} ~\sim N(X_{i,t}\beta_{i},\ \sigma^2I_{t_i})$, denoting, $\boldsymbol{\mu}_{i,t} = \textbf{X}_{i,t}\boldsymbol{\beta_i} $, then we have
$$\pi(\textbf{Y}_{i,t}|\boldsymbol{\mu}_{i,t}, \sigma^2, \boldsymbol{\beta}, \boldsymbol{\Sigma}^{-1}) \propto \frac{1}{\sqrt{\sigma^2}}exp{\{-\frac{1}{2}(\textbf{Y}_{i,t} - \boldsymbol{\mu}_{i,t})^T(\sigma^2I_{t_i})^{-1}(\textbf{Y}_{i,t} - \boldsymbol{\mu}_{i,t}) \}},$$
with the prior information given for $\boldsymbol{\beta_i} \sim N(\boldsymbol{\beta}, \Sigma^{-1})$, the posterior distribution of $\boldsymbol{\beta}_i$ is
  
  According to the knowledge of Multivariate Linear Regression Model, we know that $Y_{i,t} ~\sim N(X_{i,t}\beta_{i},\ \sigma^2I_{t_i})$, denoting, $\boldsymbol{\mu}_{i,t} = \textbf{X}_{i,t}\boldsymbol{\beta_i} $, then we have
$$\pi(\textbf{Y}_{i,t}|\boldsymbol{\mu}_{i,t}, \sigma^2, \boldsymbol{\beta}, \boldsymbol{\Sigma}^{-1}) \propto \frac{1}{\sqrt{\sigma^2}}exp{\{-\frac{1}{2}(\textbf{Y}_{i,t} - \boldsymbol{\mu}_{i,t})^T(\sigma^2I_{t_i})^{-1}(\textbf{Y}_{i,t} - \boldsymbol{\mu}_{i,t}) \}}$$
1. For $\pi(\boldsymbol{\beta_i}|.):$
$$\begin{align*}
    \pi(\boldsymbol{\beta_i}|.) &\propto f(\boldsymbol{Y_i}|\boldsymbol{\beta_i}, \sigma^2) f(\boldsymbol{\beta_i}|\boldsymbol{\beta}, \boldsymbol{\Sigma}^{-1}) \\
    &\propto \bigg(\prod^{t_i}_{t=1} \sigma^{-1}\exp\{{-\frac{(Y_{i,t} - \mu_{i,t})^2}{2\sigma^2}\}}\bigg)\bigg(\frac{1}{\sqrt{|\boldsymbol{\Sigma}|}} \exp \{ -\frac{1}{2} (\boldsymbol{\beta_i} - \boldsymbol{\beta})^T\boldsymbol{\Sigma^{-1}}(\boldsymbol{\beta_i} - \boldsymbol{\beta})\} \bigg) \\
    &\propto \exp\{-\frac{1}{2}\bigg(\sum_{t = 1}^{t_i}\sigma^{-2}(Y_{i,t} - \boldsymbol{X_{i,t}}\boldsymbol{\beta_i})^2 + (\boldsymbol{\beta_i}-\boldsymbol{\beta})^T\boldsymbol{\Sigma^{-1}}(\boldsymbol{\beta_i} - \boldsymbol{\beta})\bigg)\} \\
    &\propto  \exp\{-\frac{1}{2}\bigg((\boldsymbol{Y_{i}} - \boldsymbol{X_{i}}\boldsymbol{\beta_i})^T(\sigma^{-2}\boldsymbol{I}_{t_i \times t_i})(\boldsymbol{Y_{i}} - \boldsymbol{X_{i}}\boldsymbol{\beta_i}) + (\boldsymbol{\beta_i}-\boldsymbol{\beta})^T\boldsymbol{\Sigma}^{-1}_{8 \times 8}(\boldsymbol{\beta_i} - \boldsymbol{\beta})\bigg)\} 
\end{align*}$$

For the exponential term:
$$\begin{align*}
&\;\;\;\;(\boldsymbol{Y_{i}} - \boldsymbol{X_{i}}\boldsymbol{\beta_i})^T(\sigma^{-2}\boldsymbol{I}_{t_i \times t_i})(\boldsymbol{Y_{i}} - \boldsymbol{X_{i}}\boldsymbol{\beta_i}) + (\boldsymbol{\beta_i}-\boldsymbol{\beta})^T\boldsymbol{\Sigma}^{-1}_{8 \times 8}(\boldsymbol{\beta_i} - \boldsymbol{\beta}) \\
&  = \boldsymbol{Y_i}^T \sigma^{-2} \boldsymbol{I}\boldsymbol{Y_i}^T  + \boldsymbol{\beta_i}^T\boldsymbol{X_i}^T\sigma^{-2} \boldsymbol{I}\boldsymbol{X_i}\boldsymbol{\beta_i} - 2 \boldsymbol{Y_i}^T \sigma^{-2}\boldsymbol{I}\boldsymbol{X_i}\boldsymbol{\beta_i}  \\
& \;\;\;\;\;+ \boldsymbol{\beta_i}^T\boldsymbol{\Sigma}^{-1}\boldsymbol{\beta_i} + \boldsymbol{\beta}^T\boldsymbol{\Sigma}^{-1}\boldsymbol{\beta} - 2\boldsymbol{\beta}^T\boldsymbol{\Sigma}^{-1}\boldsymbol{\beta_i} \\
& = \boldsymbol{Y_i}^T \sigma^{-2} \boldsymbol{I}\boldsymbol{Y_i}^T + \boldsymbol{\beta}^T\boldsymbol{\Sigma}^{-1}\boldsymbol{\beta} + \boldsymbol{\beta_i}^T(\boldsymbol{\Sigma}^{-1} + \boldsymbol{X_i}^T\sigma^{-2} \boldsymbol{I}\boldsymbol{X_i})\boldsymbol{\beta_i} \\
&\;\;\;\;\;-2(\boldsymbol{Y_i}^T\sigma^{-2}\boldsymbol{I}\boldsymbol{X_i} + \boldsymbol{\beta}^T\boldsymbol{\Sigma}^{-1})\boldsymbol{\beta_i} \\
& = \boldsymbol{R} + \boldsymbol{\beta_i}^T\boldsymbol{V}\boldsymbol{\beta_i} - 2\boldsymbol{M}\boldsymbol{\beta_i}
\end{align*}$$

Where:
$$\begin{align*}
    \boldsymbol{R} &= \boldsymbol{Y_i}^T \sigma^{-2} \boldsymbol{I}\boldsymbol{Y_i}^T + \boldsymbol{\beta}^T\boldsymbol{\Sigma}^{-1}\boldsymbol{\beta}\\
    \boldsymbol{V} & = \boldsymbol{\Sigma}^{-1} + \sigma^{-2}\boldsymbol{X_i}^T\boldsymbol{X_i} \\
    \boldsymbol{M} & = \sigma^{-2}\boldsymbol{Y_i}^T\boldsymbol{X_i} + \boldsymbol{\beta}^T\boldsymbol{\Sigma}^{-1}
\end{align*}$$

Then, the exponential term can be reduced to:
$$(\boldsymbol{\beta_i} - \boldsymbol{V^{-1}M})^T\boldsymbol{V}(\boldsymbol{\beta_i} - \boldsymbol{V^{-1}M})  - \boldsymbol{M}^{T}\boldsymbol{V^{-1}}\boldsymbol{M}^{T} + \boldsymbol{R}$$

We can ignore the latter 2 term as it is not related to $\boldsymbol{\beta_i}$. That indicate:
$$\pi(\boldsymbol{\beta_i}|.) \sim N(\boldsymbol{V^{-1}M}, \boldsymbol{V^{-1}})$$

2.For $\pi(\sigma^2|.):$
$$\begin{align*}
    \pi(\sigma^2|.) & \propto f(\boldsymbol{Y}|\boldsymbol{\beta_i}, \sigma^2) \cdot \pi(\sigma^2) \\
    &\propto \bigg(\prod^n_{i=1}\prod^{t_i}_{t=1} \sigma^{-1}\exp\{{-\frac{(Y_{i,t} - \mu_{i,t})^2}{2\sigma^2}\}} \bigg)\frac{1}{\sigma^2}\\
    &\propto (\sigma^2)^{-1-\frac{\sum t_i}{2}}\prod^n_{i=1}\prod^{t_i}_{t=1} \exp\{{-\frac{(Y_{i,t} - \mu_{i,t})^2}{2\sigma^2}\}} \\
    &\propto \sigma^{-2-\sum t_i}\exp{\{-\frac{1}{2\sigma^2}\sum^n_{i=1}\sum^{ti}_{t=1}(Y_{i,t} - \mu_{i,t})^2\}} 
\end{align*}$$

So:  
$$\sigma^2 \sim \text{InvGamma}(\frac{1}{2}\sum_{i=1}^n t_i, \frac{1}{2}\sum^n_{i=1}\sum^{ti}_{t=1}(Y_{i,t} - \mu_{i,t})^2)$$

3. For $\pi(\boldsymbol{\Sigma}^{-1}|.)$:
$$\begin{align*}
    \pi(\boldsymbol{\Sigma}^{-1}|.) &\propto f(\boldsymbol{\beta_i}|\boldsymbol{\beta}, \boldsymbol{\Sigma}^{-1})f(\boldsymbol{\Sigma}^{-1}) \\
    &\propto \bigg(\prod^n_{i=1}\boldsymbol{\Sigma}^{-\frac{1}{2}} \exp\{-\frac{1}{2}(\boldsymbol{\beta_i}-\boldsymbol{\beta})^T\boldsymbol{\Sigma}^{-1}(\boldsymbol{\beta_i}-\boldsymbol{\beta}) \} \bigg) |\boldsymbol{\Sigma}|^{-(d-1)/2} \exp\{-\frac{1}{2}\boldsymbol{\Sigma}^{-1}\} \\
    &\propto |\boldsymbol{\Sigma}|^{-(n+d+1)/2} \exp\{-\frac{1}{2}tr(\boldsymbol{\Sigma^{-1}})-\frac{1}{2}\sum^n_{i=1}(\boldsymbol{\beta_i}-\boldsymbol{\beta})^T\boldsymbol{\Sigma}^{-1}(\boldsymbol{\beta_i}-\boldsymbol{\beta})\} \\
    &\propto |\boldsymbol{\Sigma}|^{-(n+d+1)/2} \exp \{ -\frac{1}{2}tr\bigg(\boldsymbol{\Sigma^{-1}}(\boldsymbol{I} + \sum_{i=1}^{n}(\boldsymbol{\beta_i}-\boldsymbol{\beta})(\boldsymbol{\beta_i}-\boldsymbol{\beta})^{T})\bigg)\}
\end{align*}$$

That indicate: 
$$\boldsymbol{\Sigma}^{-1} \sim \text{InvWhishart}\bigg(n,\boldsymbol{I} + \sum_{i=1}^{n}(\boldsymbol{\beta_i}-\boldsymbol{\beta})(\boldsymbol{\beta_i}-\boldsymbol{\beta})^{T}\bigg)$$

4. For $\pi(\boldsymbol{\beta}|.)$: 
$$\begin{align*}
    \pi(\boldsymbol{\beta}|.) &\propto f(\boldsymbol{\beta_i}|\boldsymbol{\beta}, \boldsymbol{\Sigma}^{-1})f(\boldsymbol{\beta}) \\
    &\propto \bigg(\prod^n_{i=1} \exp\{-\frac{1}{2}(\boldsymbol{\beta_i}-\boldsymbol{\beta})^T\boldsymbol{\Sigma}^{-1}(\boldsymbol{\beta_i}-\boldsymbol{\beta}) \} \bigg) \\
    &\propto \exp \{ -\frac{1}{2}\bigg(\sum_{i = 1}^{n} (\boldsymbol{\beta_i}-\boldsymbol{\beta})^T\boldsymbol{\Sigma}^{-1}(\boldsymbol{\beta_i}-\boldsymbol{\beta})\bigg)\} \\
    &\propto \exp \{-\frac{1}{2}\bigg(\sum_{i = 1}^{n} \boldsymbol{\beta_i}^T\boldsymbol{\Sigma}^{-1}\boldsymbol{\beta_i}+\boldsymbol{\beta}^Tn\boldsymbol{\Sigma}^{-1}\boldsymbol{\beta}-\sum_{i = 1}^{n}2\boldsymbol{\beta_i}^T\boldsymbol{\Sigma^{-1}}\boldsymbol{\beta}\bigg)\}
\end{align*}$$

For the exponential term, if we set: 
$$\begin{align*}
    \boldsymbol{V} &= n\boldsymbol{\Sigma}^{-1} \\
    \boldsymbol{R} &= \sum_{i = 1}^{n} \boldsymbol{\beta_i}^T\boldsymbol{\Sigma}^{-1}\boldsymbol{\beta_i} \\
    \boldsymbol{M} &= \sum_{i = 1}^{n}\bigg(\boldsymbol{\Sigma}^{-1}\boldsymbol{\beta_i}\bigg) 
\end{align*}$$

Then, use the same technique when generating $\boldsymbol{\beta_i}$
$$R + \boldsymbol{\beta}\boldsymbol{V}
\boldsymbol{\beta} - 2\boldsymbol{M}\boldsymbol{\beta} \propto (\boldsymbol{\beta} - \boldsymbol{V}^{-1}\boldsymbol{M})^{T}\boldsymbol{V}^{-1}(\boldsymbol{\beta} - \boldsymbol{V}^{-1}\boldsymbol{M})$$

(NOTE: This is the same as using OLS to estimate $\boldsymbol{\beta}$ using all $\boldsymbol{\beta_i}$)

That indicate: 

$$\boldsymbol{\beta} \sim N \bigg(\frac{1}{n}\sum_{i=1}^n \boldsymbol{\beta_i}, \frac{1}{n}\boldsymbol{\Sigma}\bigg)$$

### Markov Chain Monte Carlo

## Results and Discussions


Gibbs sampling

```{r, include = F}
library(tidyverse)
library(lubridate)
library(matrixcalc)
library(truncnorm)
library(mvtnorm)
library(bayesm)
```

```{r}
# data cleaning
df = read.csv("./hurrican356.csv") %>% 
   janitor::clean_names() %>% 
   rename(year = season) %>% 
   separate(time, into = c("date", "hour"), sep = " ") %>% 
   mutate(
     date = str_remove(date, "\\("),
     hour = str_remove(hour, "\\)")
     ) %>% 
   mutate(month = str_match(date, "-\\s*(.*?)\\s*-")) %>% 
   mutate(month = gsub("[-]", "", month)) %>% 
   filter(hour == "00:00:00" | hour == "06:00:00" | hour == "12:00:00" | hour == "18:00:00") %>% 
   mutate(
     hour = str_replace(hour, ":00:00", ""),
     hour = as.numeric(hour),
     month = month[,1],
     month = as.factor(month)
   ) %>% 
   group_by(id) %>% 
   mutate(
     delta1 = c(NA, diff(latitude)),
     delta2 = c(NA, diff(longitude)),
     delta3 = c(NA, diff(wind_kt))
   ) %>% 
   ungroup() %>% 
   na.omit() %>% 
   select(id, month, year, nature, delta1, delta2, delta3, latitude, longitude, wind_kt)
```

```{r}
test_df = df %>% filter(id == "ALLISON.1989")
```

```{r}
gibbs_fxn = function(y, niter = 20, betaistart = rep(1, 8), betastart = 1, sigstart = 0, sig2start = 0) {
  n = dim(y)[1]
  betai_m = matrix(NA, niter, 8)
  beta_m = rep(NA, niter)
  sig_m = rep(NA, niter)
  sig2_m = rep(NA, niter)
  
  betai_m[1,] = betaistart
  beta_m[1] = betastart
  sig_m[1] = sigstart
  sig2_m[1] = sig2start
  
  for(i in 2:niter){
    beta_m[i] = rnorm(1, mean = 1/n * sum(betai_m[i-1,]), sd = 1/n * sig_m[i-1])
    sig_m = rwishart(n,diag(sum((betai_m[i-1] - beta_m[i-1])*t(betai_m[i-1,] - beta_m[i-1]))))$IW
  }
}
```

```{r, message=FALSE, warning=FALSE}

library(MASS)
library(tidyverse)
library(extraDistr)
library(matrixsampling)

data = read.csv("hurrican356.csv")
```

## MCMC Code Need re-check and run

```{r, message=FALSE, warning=FALSE}

id = data %>% group_by(Season, ID) %>% summarize(n = n())
pre.data = data %>%
  select(Wind.kt, X, Month, Season, Nature, Latitude, Longitude) %>%
  mutate(Month = as.numeric(as.factor(Month)),
         Nature = as.numeric(as.factor(Nature)))

#---------Dataset Preparation----------
pre1 = function(df){
  delta.lat = sapply(2: nrow(df), function(ii){df[ii, 6] - df[ii-1, 6]})
  delta.lon = sapply(2: nrow(df), function(iii){df[iii,7] - df[iii-1, 7]})
  delta.ws = sapply(2: nrow(df), function(iv){df[iv,1] - df[iv-1, 1]})
  speed = df[-nrow(df),1]
  res = cbind(df[-1,-c(6,7)], speed, delta.lat, delta.lon, delta.ws)
  return(res)}
pre = function(df, id){
  dat = vector("list", length = nrow(id))
  id.n = cumsum(c(0, id$n))
  for (i in 1:nrow(id)){
    dat[[i]] = df[seq(id.n[i]+1, id.n[i+1]),]}
  for (i in 1:length(dat)){
    df = dat[[i]]
    if (nrow(df) != 1) dat[[i]] = pre1(df)
    else dat[[i]] = cbind(df[-c(6,7)], df[1] , delta.lat = 0, delta.lon = 0, delta.ws = 0)}
  return(dat)}
dat = pre(pre.data, id)
```


```{r}

betai = function(dat, beta, sigma, big.sig){
  res = vector("list", length = length(dat))
  # Beta_i function ~ N(V^-1*M, V^-1)
  for (i in 1:length(dat)){
    x = dat[[i]][,-1]
    y = dat[[i]][,1]
    k = big.sig + sigma^(-1) * t(x) %*% x
    m = t(x) %*% y * sigma^(-1) + big.sig %*% beta
    mu = solve(k) %*% m
    sd = solve(k)
    bi = mvrnorm(1, mu = mu, Sigma = sd)
    ssr = sum((y - x %*% bi)^2) # For calculating sigma
    res[[i]] = list(bi = bi, ssr = ssr)}
  bi2 = ""
  ssr2 = ""
  for (ii in 1:length(res)){
    bi2 = rbind(bi2, res[[i]]$bi)
    ssr2 = rbind(ssr2, res[[i]]$ssr)}
  beta.i = apply(bi2[-1,], 2, as.numeric) # Final beta.i is the 365x8 matrix 
  ssr = apply(ssr2[-1,], 2, as.numeric) # SSR is the single value for calculating sigma
  return(list(beta.i = beta.i, ssr = ssr))}

sigma = function(ssr, n = 365){
  a = 11343/2 # Need re-check
  b = sum(ssr)/2
  sig = rinvgamma(1, alpha = a, beta = b) # single value 
  return(sig)}

bsig = function(n, betai, beta){
  v0 = n
  s0 = n * t(betai - beta) %*% (betai - beta)
  bsig = rinvwishart(1, nu = v0, Omega = s0) # Covariance matrix 8x8 
  return(bsig)}

beta.fun = function(betai, bigsig){
  ols = colMeans(betai)
  beta = mvrnorm(1, mu = ols, Sigma = bigsig) # vector 1x8
  return(beta)}

mcmc = function(dat, ini.beta, ini.bsig, ini.sig, niter = 1000){
  beta.i = vector("list", length = niter + 1) # in each iteration, betai - 365x8
  sigma = rep(NA, niter)
  b.sig = vector("list", length = niter)
  beta = matrix(NA, nrow = niter, ncol = 8)
  beta.i[[1]] = betai(dat, beta = ini.beta, sigma = ini.sig, big.sig = ini.bsig)$beta.i
  for (i in 1:niter){
    sigma[i] = sigma(ssr = beta.i[[i]]$ssr)
    b.sig[i] = bsig(n = 365, betai = beta.i[[i]]$beta.i, ini.beta)
    beta[i] = beta.fun(betai = beta.i[[i]]$beta.i, bigsig = b.sig[i])
    beta.i[i + 1] = betai(dat, beta = beta[i], sigma = sigma[i], big.sig = b.sig[i])$beta.i}
  return(list(beta.i = beta.i, sigma - sigma, b.sig = b.sig, beta = beta))}
